---
title: ""
author: "Akshi Saxena"
date: "2023-01-30"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Part A

Importing a data into R, tidying it and performing a simple meaningful visualization.

Chocolate Bar Ratings is the dataset used - https://www.kaggle.com/datasets/rtatman/chocolate-bar-ratings?select=flavors_of_cacao.csv

Variables:
1. Company (Maker-if known) - Name of the company manufacturing the bar.

2. Specific Bean Originor Bar Name - The specific geo-region of origin for the bar.

3. REF - A value linked to when the review was entered in the database. Higher = more recent.

4. ReviewDate - Date of publication of the review.

5. CocoaPercent - Cocoa percentage (darkness) of the chocolate bar being reviewed.

6. CompanyLocation - Manufacturer base country.

7. Rating - Expert rating for the bar.
Rating System:
5= Elite (Transcending beyond the ordinary limits)
4= Premium (Superior flavor development, character and style)
3= Satisfactory(3.0) to praiseworthy(3.75) (well made with special qualities)
2= Disappointing (Passable but contains at least one significant flaw)
1= Unpleasant (mostly unpalatable)

8. BeanType - The variety (breed) of bean used, if provided.

9. Broad BeanOrigin - The broad geo-region of origin for the bean.


Preprocessing:
1. Renamed columns to remove white space and shorter names
2. Changed data type - cocoa_percent was string have "%" symbol. Removed the symbol and converted it to numeric.

```{r}
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
```

```{r}
setwd('..')
dir <- getwd()
```


```{r}
path <- paste(dir, "flavors_of_cacao.csv", sep="/")
cocoa_data <- read_csv(file=path)
head(cocoa_data)
```

```{r}
colnames(cocoa_data) <- c("company", "bean_orig", "ref", "review_year", "cocoa_perc", "company_loc", "rating", "bean_type", "broad_bean_orig")

cocoa_data$cocoa_perc <- gsub("%", "", as.character(cocoa_data$cocoa_perc))
cocoa_data <- transform(cocoa_data, cocoa_perc = as.numeric(cocoa_perc))

head(cocoa_data)

```

How the cocoa percentage of chocolate bars change over time? How does that affect ratings?

```{r}
ggplot(cocoa_data, aes(x= review_year, y = rating, color = cocoa_perc)) + 
    geom_point() + 
    geom_jitter() +
    geom_smooth()
```

There are more reviews each year. It looks like chocolate bars with very high cocoa percents tend to get lower ratings.

## Part B

Used data on NCAA student-athlete academic performance. The files include the codebook and tab-delimited data for team-level
Academic Progress Rates (APRs) of Division I student-athletes from 2003-2014.

```{r}

path <- paste(dir, "NCAA-D1-APR-2003-14/DS0001/26801-0001-Data.tsv", sep="/")
apr_df_raw <- read_tsv(path, na="-99")
head(apr_df_raw)
```


1. Visualizing the distributions of APRs over time.

```{r}
apr_df <- apr_df_raw %>%
  pivot_longer(cols=starts_with("APR_RATE"), names_to="YEAR", values_to="APR") %>%
  select(SCL_UNITID, SCL_NAME, SPORT_CODE, SPORT_NAME, YEAR, APR) %>%
  mutate(YEAR=as.numeric(stringr::str_sub(YEAR, start=10, 13)))
head(apr_df)
```

```{r}
ggplot(apr_df) + geom_boxplot(aes(x=as.factor(YEAR), y=APR))
```

It looks like APR is increasing over time from 2004 to 2014.


2. Visualizing the distribution of APR over time broken down by gender division:

```{r}
gender_df <- apr_df %>% filter(SPORT_CODE != 38)
gender_df$GENDER <- ifelse(gender_df$SPORT_CODE < 19, "Men", "Women")   
head(gender_df)
```

```{r}
ggplot(gender_df) + geom_boxplot(aes(x=as.factor(YEAR), y=APR, color=GENDER))
```

It seems that over the years 2004 to 2014, women’s teams have a more APR than men’s teams on an average.


3. Visualizing the distribution of APR for both men’s and women’s teams for each sport:

```{r}
df <- gender_df %>% mutate(SPORT_NAME = stringr::str_remove(SPORT_NAME, "Men's")) %>% mutate(SPORT_NAME = stringr::str_remove(SPORT_NAME, "Women's"))
ggplot(df) + geom_boxplot(aes(x=as.factor(SPORT_NAME), y=APR, color=GENDER)) + coord_flip()
```

The sports - Voleyball and Fencing have similar APR for Men and Women.
